Comparing ChatGPT-4 to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6942 Deaths in Sierra Leone from 2019-2022

### 2024-01-26 manuscript/manuscript.tex
Add model param section to methods

ChatGPT-3.5 \cite{brownLanguageModelsAre2020} and ChatGPT-4 \cite{openaiGPT4TechnicalReport2023a} are LLMs that utilize deep neural networks with transformer architectures \cite{vaswaniAttentionAllYou2017} and reinforcement learning from human feedback \cite{ouyangTrainingLanguageModels2022,christianoDeepReinforcementLearning2017,stiennonLearningSummarizeHuman2020,wirthSurveyPreferencebasedReinforcement2017} to follow instructions from prompts and provide human-level responses, with known differences in ChatGPT-4 possessing multimodal capabilites (e.g. image/voice input/output), more recent training data, and improved responses compared to ChatGPT-3 \cite{wuBriefOverviewChatGPT2023}. ChatGPT-3.5 and ChatGPT-4 were instructed with text prompts to assign CODs based on the open narrative. InterVA-5 and InSilicoVA are widely used and studied \cite{murrayUsingVerbalAutopsy2014,otiVerbalAutopsyInterpretation2010,benaraEvaluationMethodsAssigning2023,chandramohanEstimatingCausesDeath2021,tungaVerbalAutopsyModels2021,leitaoComparisonPhysiciancertifiedVerbal2014,jhaAutomatedPhysicianAssignment2019} standard statistical models for COD assignment in VAs under the openVA framework \cite{liOpenVAToolkitVerbal2023}. InterVA-5 applies Bayesian probabilistic modelling \cite{bayesEssaySolvingProblem1958} using a set of standardized symptoms from reports and related conditional probabilities from medical experts to assign CODs based on the highest probability \cite{byassIntegratedApproachProcessing2019,byassStrengtheningStandardisedInterpretation2012}. InSilicoVA improves upon InterVA (e.g. comparable probabilities across individuals, measures of uncertainty, and inclusion of additional data sources) with a hierarchical Bayesian framework and Markov Chain Monte Carlo (MCMC) simulations \cite{brooksMarkovChainMonte1998,chibMarkovChainMonte2001,hanMarkovChainMonte2001} to incorporate multiple sources of uncertainty for assigning CODs based on the highest probability \cite{mccormickProbabilisticCauseofDeathAssignment2016b}. For assigning CODs, ChatGPT-3.5 and ChatGPT-4 require prompts containing conversation-like textual instructions as input, while InterVA-5 and InSilicoVA require structured symptom and associated sociodemographic data as input.

\subsubsection{Input Data and Model Parameters}\label{methods-models-input}

For ChatGPT-3.5 and ChatGPT-4, 6939 text prompts were generated for each physician agreed record as input to instruct the models to assign CODs based on the open narratives. Parameter \verb|t_prob| was set to 0.3. Prompts were generated with a template involving \verb|<narrative>| from the free-text open narratives, and associated \verb|<age>|, (age) \verb|<unit>|, \verb|<sex>| from the standardized questionnaire:

@@

For InterVA-5 and InSilicoVA, the standardized questionnaire data from the HEAL-SL EVA were first converted into 2016 World Health Organization (WHO) VA questionnaire revision 1.5.1 Open Data Kit (ODK) format \cite{worldhealthorganizationODKVerbalAutopsy2022,nafundiODKCollectData2023} consisting of 526 variables \cite{dipasqualeReleaseODK20162016}, followed by further conversion into OpenVA format \cite{liOpenVAToolkitVerbal2023} consisting of 353 variables \cite{byassInterVA5UserGuide2020} using the \verb|pyCrossVA| Python package \cite{thomasPycrossvaPrepareData}. The \verb|Nsim| (number of iterations to run) parameter for InSilicoVA was set to 9500 (the default of 10,000 resulted in errors and ran successfully until 9500 iterations), while the \verb|HIV| (level of prevalence of human immunodeficiency virus) and \verb|Malaria| (level of prevalence of Malaria) parameters were both set to \verb|'h'| (high). The 6939 records were all converted into OpenVA formatted records for InterVA-5 and InSilicoVA.

@@

>>>>

ChatGPT-3.5 \cite{brownLanguageModelsAre2020} and ChatGPT-4 \cite{openaiGPT4TechnicalReport2023} are LLMs that utilize deep neural networks with transformer architectures \cite{vaswaniAttentionAllYou2017} and reinforcement learning from human feedback \cite{ouyangTrainingLanguageModels2022,christianoDeepReinforcementLearning2017,stiennonLearningSummarizeHuman2020,wirthSurveyPreferencebasedReinforcement2017} to follow instructions from prompts and provide human-level responses, with known differences in ChatGPT-4 possessing multimodal capabilites (e.g. image/voice input/output), more recent training data, and improved responses compared to ChatGPT-3 \cite{wuBriefOverviewChatGPT2023}. ChatGPT-3.5 and ChatGPT-4 were instructed with text prompts to assign CODs based on the open narrative. InterVA-5 and InSilicoVA are widely used and studied \cite{murrayUsingVerbalAutopsy2014,otiVerbalAutopsyInterpretation2010,benaraEvaluationMethodsAssigning2023,chandramohanEstimatingCausesDeath2021,tungaVerbalAutopsyModels2021,leitaoComparisonPhysiciancertifiedVerbal2014,jhaAutomatedPhysicianAssignment2019} standard statistical models for COD assignment in VAs under the openVA framework \cite{liOpenVAToolkitVerbal2023}. InterVA-5 applies Bayesian probabilistic modelling \cite{bayesEssaySolvingProblem1958} using a set of standardized symptoms from reports and related conditional probabilities from medical experts to assign CODs based on the highest probability \cite{byassIntegratedApproachProcessing2019,byassStrengtheningStandardisedInterpretation2012}. InSilicoVA improves upon InterVA (e.g. comparable probabilities across individuals, measures of uncertainty, and inclusion of additional data sources) with a hierarchical Bayesian framework and Markov Chain Monte Carlo (MCMC) simulations \cite{brooksMarkovChainMonte1998,chibMarkovChainMonte2001,hanMarkovChainMonte2001} to incorporate multiple sources of uncertainty for assigning CODs based on the highest probability \cite{mccormickProbabilisticCauseofDeathAssignment2016b}. For assigning CODs, ChatGPT-3.5 and ChatGPT-4 require prompts containing conversation-like textual instructions as input, while InterVA-5 and InSilicoVA require structured symptom and associated sociodemographic data as input.

\subsubsection{Input Data and Preprocessing}\label{methods-models-input}

For ChatGPT-3.5 and ChatGPT-4, 6939 text prompts were generated for each physician agreed record as input to instruct the models to assign CODs based on the open narratives. Prompts were generated with a template involving \verb|<narrative>| from the free-text open narratives, and associated \verb|<age>|, (age) \verb|<unit>|, \verb|<sex>| from the standardized questionnaire:

@@

\noindent For InterVA-5 and InSilicoVA, the standardized questionnaire data from the HEAL-SL EVA were first converted into 2016 World Health Organization (WHO) VA questionnaire revision 1.5.1 Open Data Kit (ODK) format \cite{worldhealthorganizationODKVerbalAutopsy2022,nafundiODKCollectData2023} consisting of 526 variables \cite{dipasqualeReleaseODK20162016}, followed by further conversion into OpenVA format \cite{liOpenVAToolkitVerbal2023} consisting of 353 variables \cite{byassInterVA5UserGuide2020} using the \verb|pyCrossVA| Python package \cite{thomasPycrossvaPrepareData}. The 6939 records were all converted into OpenVA formatted records for InterVA-5 and InSilicoVA.

\subsubsection{Model Parameters}\label{methods-models-params}

The parameter \verb|top_p| for ChatGPT-3.5 and ChatGPT-4, representing the sampling temperature, was set to 0.3 to consider only the top 30\% probability mass from the token results \cite{openaiOpenAIPlatformAPI2024}. Higher values closer to 1 may make the results more random, while lower values closer to 0 make the results more focused. The \verb|Nsim| (number of iterations to run) parameter \cite{liInSilicoVAProbabilisticVerbal2022} for InSilicoVA was set to 9500\footnote{The default value of Nsim=10000 for InSilicoVA ran until 9500 iterations before it stopped due to errors, thus Nsim=9500 was used and ran successfully for all iterations.}, while the \verb|HIV| (level of prevalence of human immunodeficiency virus) and \verb|Malaria| (level of prevalence of Malaria) parameters for InterVA-5 were both set to \verb|'h'| (high) \cite{thomasInterVA5ReplicateAnalyse2021}.

@@

Did not explore effects of model parameters outside the default (e.g. ChatGPT temperature, etc).
